---
output: html_document
title: "UKB results for ipgs proof-of-concept study"
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = F, message = F,warning = F, 
                      dpi = 150, fig.path = "../output/manuscript/")
suppressMessages(silent <- lapply(
  c("knitr", "kableExtra", "tidyverse", "patchwork"), 
  library, character.only = TRUE))
theme_set(theme_bw())
```

```{r load-data}
ukb_subset_names <- c("training", "validation", "testing")
ukb_subset_names_clean <- c("Training", "Validation", "Testing")
ukb_subsets <- map(set_names(ukb_subset_names), function(set) {
  read_csv(paste0("../data/processed/ukb_", set, "_set.csv"), show_col_types = FALSE) %>%
    mutate(across(matches("^gPC"), ~ . * bmi, .names = "bmiBy{.col}"),
           across(matches("^gPC"), ~ . * whr, .names = "whrBy{.col}"))
})
```

# Project summary

```{r workflow-diagram}
knitr::include_graphics("../doc/workflow_diagram.pdf")
```

# Data distributions

Plots reflect the testing set.

```{r configuration}
exposures <- c("bmi")
exposures_clean <- c("BMI")

outcomes <- c("hscrp_log", "hba1c", "ldl_statinadj", "tg_log", "alt_log", 
              "ggt_log", "hdl", "alb", "lipA_log", "vitD", "apoA", 
              "apoB_statinadj", "ast_log", "chol_statinadj", "creatinine", 
              "cysC_log", "bilirubin_dir_log", "glu", "shbg", "bilirubin_tot_log")
outcomes_clean <- c("log(hsCRP)", "HbA1c", "LDL-C", "log(TG)", "log(ALT)",
                    "log(GGT)", "HDL-C", "Albumin", "log(LipA)", "Vitamin D",
                    "ApoA", "ApoB", "log(AST)", "Cholesterol", "Creatinine",
                    "log(cystatin C)", "log(Bilirubin-Direct)", "Glucose", 
                    "SHBG", "log(Bilirubin-Total")

threshold_vec <- 5 * 10 ** seq(-8, -2)
threshold_names <- paste0("Pt_", c("5e-08", "5e-07", "5e-06", "5e-05", "0.0005", 
                                   "0.005", "0.05"))
threshold_names_clean <- c("P<5e-8", "P<5e-7", "P<5e-6", "P<5e-5", "P<5e-4", 
                           "P<0.005", "P<0.05")

pgs_types <- c("mpgs", "ipgs", "vpgs")
pgs_types_clean <- c("mPGS", "iPGS", "vPGS")

pgs_config_df <- expand_grid(
  e = exposures,
  y = outcomes,
  pgs_type = pgs_types,
  set = names(ukb_subsets),
  thresh_name = threshold_names
) %>%
  mutate(tag = case_when(
    pgs_type == "mpgs" ~ paste0(y),
    pgs_type == "ipgs" ~ paste0(e, "_", y),
    pgs_type == "vpgs" ~ paste0(y, "_vQTL")
  ))
```

# PGS performance

```{r pgs-testing-prep}
read_pgs <- function(tag) {
  pgs_df <- read_csv(paste0("../data/processed/pgs/", tag, "_all_pgs.csv"), 
                     show_col_types = FALSE)
  pgs_df
}

read_pgs_prsice <- function(tag) {
  pgs_df <- read_delim(paste0("../data/processed/pgs/", tag, ".all_score"), 
                       delim = " ", show_col_types = FALSE) %>%
    select(-FID, id = IID, everything())
  pgs_df
}

test_pgs_main <- function(y, tag, set, thresh_name, covars = NULL) {
  pgs_df <- read_pgs_prsice(tag)
  if (thresh_name == "stacked") {
    stack_weights <- pull(filter(stack_weights_df, tag == .env$tag), stack_weights)[[1]]
    pgs_df$stacked <- calc_stacked_pgs(pgs_df, stack_weights)
  }
  pgs_df <- select(pgs_df, id, pgs = {{thresh_name}})
  regression_df <- inner_join(ukb_subsets[[set]], pgs_df, by = "id")
  lm_form_str <- paste0(y, " ~ pgs")
  if (!is.null(covars)) {
    lm_form_str <- paste0(lm_form_str, " + ", paste(covars, collapse = " + "))
  }
  tryCatch({
    lm_fit <- lm(as.formula(lm_form_str), data = regression_df)
    lm_fit %>%
      broom::tidy() %>%
      filter(term == "pgs")
  }, error = function(e) tibble())
}

test_pgs_gxe <- function(e, y, tag, set, thresh_name, covars = NULL) {
  pgs_df <- tryCatch({
    pgs_df <- read_pgs_prsice(tag)
    pgs_df <- select(pgs_df, id, pgs = {{thresh_name}})
    regression_df <- inner_join(ukb_subsets[[set]], pgs_df, by = "id")
  }, error = function(e) tibble())
  lm_form_str <- paste0(y, " ~ pgs * ", e)
  if (!is.null(covars)) {
    lm_form_str <- paste0(lm_form_str, " + ", paste(covars, collapse = " + "))
  }
  tryCatch({
    lm_fit <- lm(as.formula(lm_form_str), data = regression_df)
    lm_fit %>%
      broom::tidy() %>%
      filter(term == paste0("pgs:", e))
  }, error = function(e) tibble())
}

gPCs <- paste0("gPC", 1:10)
primary_covars <- c("sex", "age", "age_squared", "ageBySex", gPCs)
```

## Positive control: GWAS and PGS main effects

```{r pgs-main-results}
pgs_test_main_res_df <- readRDS("../data/processed/manuscript/pgs_test_main_res_df.rds")
optimal_pgs_main_df <- readRDS("../data/processed/manuscript/optimal_pgs_main_df.rds")
```

We first show the results from a "standard" PGS pipeline, using summary statistics from GWAS and evaluating performance based on the PGS main effect. This acts as a positive control and illustrates the flow of the pipeline across training, optimization, and testing sets.

```{r plot-pgs-main-results, fig.asp=0.5}
pgs_test_main_plt_df <- pgs_test_main_res_df %>%
  left_join(optimal_pgs_main_df, by = c("y", "pgs_type", "thresh_name")) 
pgs_test_main_plt_df %>%
  filter(y == "hscrp_log") %>%
  mutate(color = if_else(set == "validation", optimal, NA),
         statistic = if_else(set == "testing" & optimal == FALSE, NA, statistic),
         y = factor(y, levels = outcomes, labels = outcomes_clean),
         set = factor(set, levels = ukb_subset_names, 
                      labels = ukb_subset_names_clean),
         thresh_name = factor(thresh_name, 
                              levels = threshold_names,
                              labels = threshold_names_clean)) %>%
  ggplot(aes(x = thresh_name, y = statistic, color = color)) +
  geom_bar(stat = "identity") +
  scale_color_discrete(name = "Optimal", na.translate = FALSE) +
  facet_wrap(vars(set), nrow = 1, ncol = 3, scale = "free_y") +
  labs(x = "", y = "PGS main effect z-statistic",
       title = "Example for log(hsCRP)") +
  theme(axis.text.x = element_text(angle = 25, hjust = 0.9))

pgs_test_main_plt_df %>%
  filter(set == "testing",
         optimal == TRUE) %>%
  mutate(Biomarker = outcomes_clean[match(y, outcomes)],
         z = round(statistic, 1)) %>%
  select(Biomarker, optimal_thres = thresh_name, z) %>%
  arrange(desc(z)) %>%
  kable(caption = "mPGS main effects for all biomarkers") %>%
  kable_styling(full_width = FALSE)
```

## PGS optimization for PGSxE

```{r mpgs-results}
mpgs_test_res_df <- readRDS("../data/processed/manuscript/mpgs_test_res_df.rds")
optimal_mpgs_df <- readRDS("../data/processed/manuscript/optimal_mpgs_df.rds")
```

```{r plot-mpgs-results}
mpgs_test_res_df %>%
  filter(set == "validation") %>%
  left_join(optimal_mpgs_df, by = c("e", "y", "pgs_type", "thresh_name")) %>%
  mutate(y = factor(y, levels = outcomes, labels = outcomes_clean),
         thresh_name = factor(thresh_name, 
                              levels = threshold_names,
                              labels = threshold_names_clean)) %>%
  ggplot(aes(x = thresh_name, y = statistic, color = optimal)) +
  geom_bar(stat = "identity") +
  scale_color_discrete(name = "Optimal", na.translate = FALSE) +
  facet_wrap(vars(y, e), ncol = 5, scale = "free_y") +
  labs(x = "", y = "mPGS GxE z-statistic",
       title = "Optimization of mPGS for PGSxE") +
  theme(axis.text.x = element_text(angle = 25, hjust = 0.9))
```

```{r ipgs-results}
ipgs_test_res_df <- readRDS("../data/processed/manuscript/ipgs_test_res_df.rds")
optimal_ipgs_df <- readRDS("../data/processed/manuscript/optimal_ipgs_df.rds")

# ipgs_test_res_df <- pgs_config_df %>%
#   filter(pgs_type == "ipgs") %>%
#   rowwise() %>%
#   mutate(lm_fit = list(test_pgs_gxe(e, y, tag, set, thresh_name, 
#                                     c(primary_covars, paste0(e, "By", gPCs))))) %>%
#   ungroup() %>%
#   unnest(lm_fit)
# 
# # e <- "whr"
# # y <- "hba1c"
# # tag <- "whr_hba1c"
# # pheno_df <- ukb_subsets$validation
# 
# optimal_ipgs_df <- ipgs_test_res_df %>%
#   filter(set == "validation") %>%
#   group_by(e, y) %>%
#   filter(statistic == max(statistic)) %>%
#   ungroup() %>%
#   mutate(optimal = TRUE) %>%
#   select(e, y, pgs_type, thresh_name, optimal)
```

```{r plot-ipgs-results}
ipgs_test_res_df %>%
  filter(set == "validation") %>%
  left_join(optimal_ipgs_df, by = c("e", "y", "pgs_type", "thresh_name")) %>%
  mutate(y = factor(y, levels = outcomes, labels = outcomes_clean),
         thresh_name = factor(thresh_name, 
                              levels = threshold_names,
                              labels = threshold_names_clean)) %>%
  ggplot(aes(x = thresh_name, y = statistic, color = optimal)) +
  geom_bar(stat = "identity") +
  scale_color_discrete(name = "Optimal", na.translate = FALSE) +
  facet_wrap(vars(y), ncol = 5, scale = "free_y") +
  labs(x = "", y = "iPGS GxE z-statistic",
       title = "Optimization of iPGS for PGSxBMI") +
  theme(axis.text.x = element_text(angle = 25, hjust = 0.9))
```

```{r vpgs-results}
vpgs_test_res_df <- readRDS("../data/processed/manuscript/vpgs_test_res_df.rds")
optimal_vpgs_df <- readRDS("../data/processed/manuscript/optimal_vpgs_df.rds")

# vpgs_test_res_df <- pgs_config_df %>%
#   filter(pgs_type == "vpgs") %>%
#   rowwise() %>%
#   mutate(lm_fit = list(test_pgs_gxe(e, y, tag, set, thresh_name, 
#                                     c(primary_covars, paste0(e, "By", gPCs))))) %>%
#   ungroup() %>%
#   unnest(lm_fit)
# 
# optimal_vpgs_df <- vpgs_test_res_df %>%
#   filter(set == "validation") %>%
#   group_by(e, y) %>%
#   filter(statistic == max(statistic)) %>%
#   ungroup() %>%
#   mutate(optimal = TRUE) %>%
#   select(e, y, pgs_type, thresh_name, optimal)
```

```{r plot-vpgs-results}
vpgs_test_res_df %>%
  filter(set == "validation") %>%
  left_join(optimal_vpgs_df, by = c("e", "y", "pgs_type", "thresh_name")) %>%
  mutate(y = factor(y, levels = outcomes, labels = outcomes_clean),
         thresh_name = factor(thresh_name, 
                              levels = threshold_names,
                              labels = threshold_names_clean)) %>%
  ggplot(aes(x = thresh_name, y = statistic, color = optimal)) +
  geom_bar(stat = "identity") +
  scale_color_discrete(name = "Optimal", na.translate = FALSE) +
  facet_wrap(vars(y), ncol = 5, scale = "free_y") +
  labs(x = "", y = "vPGS GxE z-statistic",
       title = "Optimization of vPGS for PGSxBMI") +
  theme(axis.text.x = element_text(angle = 25, hjust = 0.9))
```

## Testing of optimized PGS

```{r plot-pgs-results, fig.asp=1.2}
all_pgs_test_res_df <- bind_rows(
  inner_join(mpgs_test_res_df, optimal_mpgs_df),
  inner_join(ipgs_test_res_df, optimal_ipgs_df),
  inner_join(vpgs_test_res_df, optimal_vpgs_df)
)

n_biomarkers <- length(outcomes)
bonferroni <- 0.05 / n_biomarkers

outcomes_ordered <- all_pgs_test_res_df %>%
  filter(set == "testing") %>%
  arrange(p.value) %>%
  distinct(y) %>%
  pull(y)

all_pgs_test_res_df %>%
  filter(set == "testing") %>%
  mutate(e = factor(e, levels = exposures, labels = exposures_clean),
         y = factor(y, levels = outcomes_ordered),
         y = fct_relabel(y, ~ outcomes_clean[match(., outcomes)]),
         pgs_type = factor(pgs_type, levels = pgs_types, labels = pgs_types_clean),
         set = factor(set, levels = names(ukb_subsets), 
                      labels = paste0(str_to_title(names(ukb_subsets)), " set")),
         thresh_name = factor(thresh_name, 
                              levels = threshold_names,
                              labels = threshold_names_clean)) %>%
  ggplot(aes(x = pgs_type, y = statistic, fill = pgs_type)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, color = "gray") +
  geom_hline(yintercept = qnorm(0.05 / 2, lower.tail = FALSE), color = "gray", linetype = "dashed") +
  geom_hline(yintercept = qnorm(bonferroni / 2, lower.tail = FALSE), linetype = "dashed") +
  scale_fill_discrete(name = "PGS type") +
  facet_wrap(vars(y), ncol = 4, scale = "fixed") +
  labs(x = "", y = "Interaction test z-statistic") +
  theme(axis.text.x = element_text(angle = 20, hjust = 0.9))
```

```{r plot-pgs-results-2, fig.asp=0.6}
all_pgs_test_res_df %>%
  filter(set == "testing") %>%
  mutate(e = factor(e, levels = exposures, labels = exposures_clean),
         y = factor(y, levels = outcomes_ordered),
         y = fct_relabel(y, ~ outcomes_clean[match(., outcomes)]),
         pgs_type = factor(pgs_type, levels = pgs_types, labels = pgs_types_clean),
         set = factor(set, levels = names(ukb_subsets), 
                      labels = paste0(str_to_title(names(ukb_subsets)), " set")),
         thresh_name = factor(thresh_name, 
                              levels = threshold_names,
                              labels = threshold_names_clean),
         l95 = estimate - 1.96 * std.error,
         u95 = estimate + 1.96 * std.error) %>%
  ggplot(aes(x = pgs_type, y = estimate, color = pgs_type)) +
  # geom_bar(stat = "identity") +
  geom_point() +
  geom_errorbar(aes(ymin = l95, ymax = u95), width = 0.2) +
  geom_hline(yintercept = 0, color = "gray") +
  scale_fill_discrete(name = "PGS type") +
  facet_wrap(vars(y), ncol = 4, scale = "fixed") +
  labs(x = "", y = "Interaction effect estimate (95% CI)",
       title = "PGSxE testing (testing set)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 0.9))

all_pgs_test_res_df %>%
  filter(set == "testing") %>%
  mutate(e = factor(e, levels = exposures, labels = exposures_clean),
         y = factor(y, levels = outcomes_ordered),
         y = fct_relabel(y, ~ outcomes_clean[match(., outcomes)]),
         pgs_type = factor(pgs_type, levels = pgs_types, labels = pgs_types_clean),
         set = factor(set, levels = names(ukb_subsets), 
                      labels = paste0(str_to_title(names(ukb_subsets)), " set")),
         thresh_name = factor(thresh_name, 
                              levels = threshold_names,
                              labels = threshold_names_clean),
         l95 = estimate - 1.96 * std.error,
         u95 = estimate + 1.96 * std.error) %>%
  ggplot(aes(x = y, y = estimate, color = pgs_type)) +
  geom_hline(yintercept = 0, color = "gray") +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = l95, ymax = u95), width = 0.5,
                position = position_dodge(width = 0.5)) +
  scale_fill_discrete(name = "PGS type") +
  # facet_wrap(vars(y), ncol = 4, scale = "fixed") +
  labs(x = "", y = "Interaction effect estimate (95% CI)",
       title = "PGSxE testing (testing set)") +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.9))
```

```{r expanded-pgs-results}
bm_level_summary_df <- all_pgs_test_res_df %>%
  filter(set == "testing") %>%
  group_by(y) %>%
  summarise(any_nom = any(p.value < 0.05),
            any_bonf = any(p.value < bonferroni))

all_pgs_test_res_df %>%
  filter(set == "testing",
         p.value < bonferroni) %>%
  group_by(y) %>%
  summarise(top_pgs_type = pgs_type[which.max(statistic)],
            .groups = "drop") %>%
  group_by(top_pgs_type) %>%
  summarise(n = n(),
            biomarkers = paste(unique(y), collapse = ", "),
            .groups = "drop") %>%
  arrange(desc(n)) %>%
  select(`Top PGS type` = top_pgs_type, N = n, Biomarkers = biomarkers) %>%
  kable(caption = "Top-performing PGS type per biomarker (of those with any Bonferroni significant PGS)") %>%
  kable_styling(full_width = FALSE)

all_pgs_test_res_df %>%
  filter(set == "testing") %>%
  arrange(desc(abs(statistic))) %>%
  select(y, pgs_type, thresh_name, estimate, statistic, p.value) %>%
  mutate(p.value = as.character(signif(p.value, 3))) %>%
  head(10) %>%
  kable(caption = "Top PGSxE tests") %>%
  kable_styling()

# bmi_y_corr_df <- expand_grid(
#   y = outcomes
# ) %>%
#   rowwise() %>%
#   mutate(corr = cor.test(ukb_subsets$testing$bmi, ukb_subsets$testing[[y]])$estimate) %>%
#   ungroup() %>%
#   select(y, corr) %>%
#   arrange(desc(corr))
# 
# all_pgs_test_res_df %>%
#   filter(set == "testing") %>%
#   inner_join(bmi_y_corr_df, by = "y") %>%
#   select(y, pgs_type, p.value, corr) %>%
#   arrange(desc(corr)) %>%
#   group_by(y) %>%
#   summarise(ipgs_better = p.value[pgs_type == "mpgs"] / p.value[pgs_type == "ipgs"],
#             corr = corr[1]) %>%
#   arrange(desc(ipgs_better))
```

Of the biomarkers tested, `r sum(bm_level_summary_df$any_nom)` had at least one nominally significant PGS interaction test (P < 0.05), and `r sum(bm_level_summary_df$any_bonf)` had at least one significant interaction test after Bonferroni correction (P < 0.05 / `r length(outcomes)``).

## Sensitivity analyses

```{r plot-sensitivity-results}
sensitivity_test_plt_df <- readRDS("../data/processed/manuscript/sensitivity_test_plt_df.rds")

robust_sensitivity_plt <- sensitivity_test_plt_df %>%
  ggplot(aes(x = statistic_primary, y = statistic_robust)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  labs(x = "Primary model z-statistic", y = "Robust SE model z-statistic")

eSq_sensitivity_plt <- sensitivity_test_plt_df %>%
  ggplot(aes(x = statistic_primary, y = statistic_eSq)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  labs(x = "Primary model z-statistic", y = "Nonlinear E model z-statistic")

bmiPGS_sensitivity_plt <- sensitivity_test_plt_df %>%
  ggplot(aes(x = statistic_primary, y = statistic_bmiPGS)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  coord_cartesian(ylim = c(NA, max(sensitivity_test_plt_df$statistic_primary))) +
  labs(x = "Primary model z-statistic", y = "BMI PGS z-statistic")

layout <- "
AABB
#CC#
"
(robust_sensitivity_plt + eSq_sensitivity_plt + bmiPGS_sensitivity_plt) +
  plot_layout(design = layout) +
  plot_annotation(tag_levels = "a")
```

## Summary plots

```{r sankey}
library(networkD3)

# Define the nodes (i.e., the stages of the process)
nodes <- data.frame(
  name = c("Total biomarkers\n(20)", 
           "Significant PGSxE (14)", "No significant PGSxE",
           "mPGS strongest (3)", "iPGS strongest (11)", #"vPGS strongest",
           "Replication in AoU (8)", "No replication in AoU (3)"),
  grp = c("source", 
          rep("significance", 2), 
          rep("strongest", 2),
          rep("AoU", 2))
)

# Define the links (i.e., the connections between stages)
links <- data.frame(
  source = c(0, 0, rep(1, 2), rep(4, 2)), # Start from nodes by their index
  target = c(1, 2, 3, 4, 5, 6), # Flow to other nodes
  value = c(14, 6, 3, 11, 8, 3)
)

# Create the Sankey diagram
sankey <- networkD3::sankeyNetwork(
  Links = links, Nodes = nodes, 
  Source = "source", Target = "target", 
  Value = "value", 
  NodeID = "name", NodeGroup = "grp",
  units = "Tests", 
  fontSize = 12, nodeWidth = 30, nodePadding = 10,
  height = 300, width = 700, sinksRight = FALSE
)

# Display the Sankey plot
sankey
```

# Specific visualizations - BMI and log(ALT)

```{r example-setup}
e <- "bmi"
y <- "alt_log"
tag <- "alt_log"
thresh_name <- "Pt_5e-08"
e_clean <- "BMI"
y_clean <- "log(ALT)"
pgs_df <- read_pgs_prsice(tag) %>%
  select(id, pgs = {{thresh_name}})
regression_df <- inner_join(ukb_subsets$testing, pgs_df, by = "id") %>%
  filter(!is.na(.data[[e]]), !is.na(.data[[y]])) %>%
  mutate(pgs_bin = cut(pgs, quantile(pgs, seq(0, 1, length.out = 11)), 
                       include.lowest = TRUE, labels = paste0("Q", 1:10)))
```

```{r example-e-y, out.width="30%"}
regression_df %>%
  ggplot(aes_string(x = e, y = y)) +
  geom_smooth(method = "gam", formula = y ~ rms::rcs(x, 5)) +
  labs(x = e_clean,
       y = y_clean)

print(paste0("Example E-Y corr: ", 
             round(cor(regression_df[[e]], regression_df[[y]]), 2)))
```

```{r example-main-effects, fig.asp=0.5}
pgs_test_main_res_df %>%
  left_join(optimal_pgs_main_df, by = c("y", "pgs_type", "thresh_name")) %>%
  filter(y == .env$y) %>%
  mutate(color = if_else(set == "validation", optimal, NA),
         statistic = if_else(set == "testing" & optimal == FALSE, NA, statistic),
         y = factor(y, levels = outcomes, labels = paste0(outcomes_clean, " set")),
         set = factor(set, levels = ukb_subset_names, 
                      labels = ukb_subset_names_clean),
         thresh_name = factor(thresh_name, 
                              levels = threshold_names,
                              labels = threshold_names_clean)) %>%
  ggplot(aes(x = thresh_name, y = statistic, color = color)) +
  geom_bar(stat = "identity") +
  scale_color_discrete(name = "Optimal", na.translate = FALSE) +
  facet_wrap(vars(set), nrow = 1, scale = "free_y") +
  labs(x = "", y = "PGS main effect z-statistic",
       title = "Positive control: standard GWAS/main effects pipeline") +
  theme(axis.text.x = element_text(angle = 25, hjust = 0.9))
```

```{r example-test, fig.width=5, fig.asp=0.5}
all_pgs_test_res_df %>%
  filter(set == "testing",
         e == .env$e,
         y == .env$y) %>%
  mutate(e = factor(e, levels = exposures, labels = exposures_clean),
         y = factor(y, levels = outcomes, labels = outcomes_clean),
         pgs_type = factor(pgs_type, levels = pgs_types, labels = pgs_types_clean),
         set = factor(set, levels = names(ukb_subsets), 
                      labels = paste0(str_to_title(names(ukb_subsets)), " set")),
         thresh_name = factor(thresh_name, 
                              levels = threshold_names,
                              labels = threshold_names_clean)) %>%
  ggplot(aes(x = pgs_type, y = statistic, fill = pgs_type)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, color = "gray") +
  scale_fill_discrete(name = "PGS type") +
  labs(x = "", y = "Interaction test z-statistic",
       title = "PGSxBMI (testing set)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 0.9))
```

```{r example-strat-plots}
quantile_lines_plt <- regression_df %>%
  ggplot(aes_string(x = e, y = y, group = "pgs_bin", color = "pgs_bin")) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", formula = "y ~ x") +
  scale_color_discrete(name = "PGS bin") +
  coord_cartesian(xlim = quantile(regression_df[[e]], c(0.01, 0.99)),
                  ylim = quantile(regression_df[[y]], c(0.01, 0.99))) +
  labs(x = e_clean,
       y = y_clean)

test_ey <- function(df, e, y, covars) {
  lm_form_str <- paste0(y, " ~ ", e, " + ", paste(covars, collapse = " + "))
  lm_fit <- lm(as.formula(lm_form_str), data = df)
  lm_fit %>%
    broom::tidy() %>%
    filter(term == e)
}

quantile_effects_df <- regression_df %>%
  nest(data = -pgs_bin) %>%
  rowwise() %>%
  mutate(ey_lm = test_ey(data, e, y, primary_covars)) %>%
  select(-data) %>%
  unnest(ey_lm)
quantile_effects_plt <- quantile_effects_df %>%
  select(pgs_bin, estimate) %>%
  ggplot(aes(x = pgs_bin, y = estimate)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "gray") +
  labs(x = "PGS bin",
       y = paste0(e_clean, " - ", y_clean, " effect size"))

quantile_lines_plt + quantile_effects_plt

quantile_effects_df %>%
  arrange(pgs_bin) %>%
  select(pgs_bin, estimate) %>%
  mutate(factor_change = estimate / estimate[1],
         across(-pgs_bin, ~ round(.x, 3))) %>%
  kable(caption = "PGS-stratified BMI-risk factor effects") %>%
  kable_styling(full_width = FALSE)
```
