---
output: html_document
title: "UKB results for ipgs proof-of-concept study"
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = F, message = F,warning = F, 
                      dpi = 150, fig.path = "../output/ukb_results/")
suppressMessages(silent <- lapply(
  c("knitr", "kableExtra", "tidyverse", "patchwork"), 
  library, character.only = TRUE))
theme_set(theme_bw())
```

```{r load-data}
ukb_subset_names <- c("training", "validation", "testing")
ukb_subsets <- map(set_names(ukb_subset_names), function(set) {
  read_csv(paste0("../data/processed/ukb_", set, "_set.csv"), show_col_types = FALSE) %>%
    mutate(across(matches("^gPC"), ~ . * bmi, .names = "bmiBy{.col}"),
           across(matches("^gPC"), ~ . * whr, .names = "whrBy{.col}"))
})
```

# Project summary

```{r workflow-diagram}
knitr::include_graphics("../doc/workflow_diagram.pdf")
```

Overall dataset consists of European-ancestry individuals from the UKBB (based on Pan-UKBB; N ~ 320,000).

* 50% training set: GWAS/GWIS
* 25% validation set: PGS parameter optimization
* 25% testing set: evaluation

Primary metric for evaluation (validation & testing): significance of regression PGSxE term

Variables being manipulated:

* Exposure: BMI, WHR
* Outcome: log(hsCRP), HbA1c
* PGS type: mPGS, iPGS (vPGS & pPGS pending)
* PGS name/threshold: 0.05, 0.01, 1e-5, 5e-8

# Data distributions

Plots reflect the testing set.

```{r histograms}
plot_histogram <- function(f) {
  ukb_subsets$testing %>%
    select(x = {{ f }}) %>%
    filter(!is.na(x)) %>%
    ggplot(aes(x = x)) +
    geom_histogram(stat = "bin") +
    labs(x = f)
}

hscrp_hist <- plot_histogram("hscrp_log")
hba1c_hist <- plot_histogram("hba1c")

hscrp_hist + hba1c_hist
```

# PGS performance

```{r configuration}
exposures <- c("bmi", "whr")
exposures_clean <- c("BMI", "WHR")

outcomes <- c("hscrp_log", "hba1c")
outcomes_clean <- c("log(hsCRP)", "HbA1c")

threshold_vec <- c("0.05", "0.01", "0.00001", "0.00000005")
thresh_names <- paste0("thresh", threshold_vec)
thresh_names_clean <- c("P<0.05", "P<0.01", "P<1e-5", "P<5e-8")

pgs_types <- c("gxe", "main")
pgs_types_clean <- c("iPGS", "mPGS")

pgs_config_df <- expand_grid(
  e = exposures,
  y = outcomes,
  pgs_type = pgs_types
) %>%
  mutate(tag = case_when(
    pgs_type == "gxe" ~ paste0(e, "_", y),
    # type == "prod" ~ paste0(e, "_", y, "_prod"),
    pgs_type == "main" ~ paste0(y),
  ))
```

```{r stacking}
get_stack_weights <- function(e, y, tag, thresh_names, pheno_df) {
  pgs_df <- read_csv(paste0("../data/processed/pgs/", tag, "_all_pgs.csv"))
  df <- pheno_df %>%
    inner_join(pgs_df, by = "id") %>%
    filter(!is.na(!! sym(e)), !is.na(!! sym(y)))
  me_lm_str <- paste0(y, " ~ ", e, " + ", paste(thresh_names, collapse = " + "))
  me_lm_fit <- lm(as.formula(me_lm_str), data = df, na.action = na.exclude)
  me_lm_resids <- resid(me_lm_fit)
  scaled_pgs_mat <- df %>%
    select(all_of(thresh_names)) %>%
    mutate(across(everything(), ~ ifelse(all(.x == 0), 0, scale(.x)))) %>%
    as.matrix()
  gxe_mat <- scaled_pgs_mat * df[[e]]  # Column-wise multiplication
  gxe_fit_cv <- glmnet::cv.glmnet(gxe_mat, me_lm_resids,
                                  alpha = 0, lower.limits = 0, nfolds = 5)
  weights <- coef(gxe_fit_cv, s = "lambda.min")[thresh_names, ]
  weights_norm <- weights / sum(weights)
  weights_norm
}

calc_stacked_pgs <- function(df, weights) {
  scaled_pgs_mat <- df %>%
    select(all_of(names(weights))) %>%
    mutate(across(everything(), scale)) %>%
    as.matrix()
  stacked_pgs <- scaled_pgs_mat %*% weights
  stacked_pgs
}

stack_weights_df <- pgs_config_df %>%
  rowwise() %>%
  mutate(stack_weights = list(get_stack_weights(e, y, tag, thresh_names, ukb_subsets$validation))) %>%
  ungroup()
```

```{r pgs-testing-prep}
# get_pgs <- function(e, y, thresh_names, tag) {
#   pgs_df <- read_csv(paste0("../data/processed/pgs/", tag, "_all_pgs.csv")) %>%
#     select(id, all_of(thresh_names))
#   stack_weights <- get_stack_weights(e, y, tag, thresh_names, ukb_subsets$validation)
#   pgs_df$stacked <- calc_stacked_pgs(pgs_df, stack_weights)
#   pgs_df
# }
test_pgs_main <- function(y, tag, set, thresh_name, covars = NULL) {
  pgs_df <- read_csv(paste0("../data/processed/pgs/", tag, "_all_pgs.csv"),
                     show_col_types = FALSE)
  if (thresh_name == "stacked") {
    stack_weights <- pull(filter(stack_weights_df, tag == .env$tag), stack_weights)[[1]]
    pgs_df$stacked <- calc_stacked_pgs(pgs_df, stack_weights)
  }
  pgs_df <- select(pgs_df, id, pgs = {{thresh_name}})
  regression_df <- inner_join(ukb_subsets[[set]], pgs_df, by = "id")
  lm_form_str <- paste0(y, " ~ pgs")
  if (!is.null(covars)) {
    lm_form_str <- paste0(lm_form_str, " + ", paste(covars, collapse = " + "))
  }
  tryCatch({
    lm_fit <- lm(as.formula(lm_form_str), data = regression_df)
    lm_fit %>%
      broom::tidy() %>%
      filter(term == "pgs")
  }, error = function(e) tibble())
}

test_pgs_gxe <- function(e, y, tag, set, thresh_name, covars = NULL) {
  pgs_df <- read_csv(paste0("../data/processed/pgs/", tag, "_all_pgs.csv"), 
                     show_col_types = FALSE)
  if (thresh_name == "stacked") {
    stack_weights <- pull(filter(stack_weights_df, tag == .env$tag), stack_weights)[[1]]
    pgs_df$stacked <- calc_stacked_pgs(pgs_df, stack_weights)
  }
  pgs_df <- select(pgs_df, id, pgs = {{thresh_name}})
  regression_df <- inner_join(ukb_subsets[[set]], pgs_df, by = "id")
  lm_form_str <- paste0(y, " ~ pgs * ", e)
  if (!is.null(covars)) {
    lm_form_str <- paste0(lm_form_str, " + ", paste(covars, collapse = " + "))
  }
  tryCatch({
    lm_fit <- lm(as.formula(lm_form_str), data = regression_df)
    lm_fit %>%
      broom::tidy() %>%
      filter(term == paste0("pgs:", e))
  }, error = function(e) tibble())
}

gPCs <- paste0("gPC", 1:10)
primary_covars <- c("sex", "age", "age_squared", "ageBySex", gPCs)
```

```{r pgs-main-results}
pgs_test_main_res_df <- pgs_config_df %>%
  filter(pgs_type == "main") %>%
  select(y, pgs_type, tag) %>%
  distinct() %>%
  cross_join(expand_grid(set = names(ukb_subsets),
                         thresh_name = thresh_names)) %>%
  rowwise() %>%
  mutate(lm_fit = list(test_pgs_main(y, tag, set, thresh_name, primary_covars))) %>%
  ungroup() %>%
  unnest(lm_fit)
```

```{r pgs-results}
pgs_test_res_df <- pgs_config_df %>%
  cross_join(expand_grid(set = names(ukb_subsets),
                         thresh_name = c(thresh_names, "stacked"))) %>%
  rowwise() %>%
  mutate(lm_fit = list(test_pgs_gxe(e, y, tag, set, thresh_name, 
                                    c(primary_covars, paste0(e, "By", gPCs))))) %>%
  ungroup() %>%
  unnest(lm_fit)

optimal_pgs_df <- pgs_test_res_df %>%
  filter(set == "validation",
         thresh_name != "stacked") %>%
  group_by(e, y, pgs_type) %>%
  filter(statistic == max(statistic)) %>%
  ungroup() %>%
  mutate(optimal = TRUE) %>%
  select(e, y, pgs_type, thresh_name, optimal)

# e <- "whr"
# y <- "hba1c"
# tag <- "whr_hba1c_prod"
# pheno_df <- ukb_subsets$validation
```

```{r plot-pgs-main-results, fig.asp=0.5}
pgs_test_main_res_df %>%
  filter(set == "testing",
         pgs_type == "main") %>%
  mutate(y = factor(y, levels = outcomes, labels = outcomes_clean),
         pgs_type = factor(pgs_type, levels = pgs_types, labels = pgs_types_clean),
         thresh_name = factor(thresh_name, 
                           levels = thresh_names,
                           labels = thresh_names_clean)) %>%
  ggplot(aes(x = thresh_name, y = statistic)) +
  geom_bar(stat = "identity") +
  facet_wrap(vars(y), ncol = 2, scale = "free_y") +
  labs(x = "", y = "PGS z-statistic (testing set)",
       title = "Positive control: standard GWAS/main effects pipeline") +
  theme(axis.text.x = element_text(angle = 20, hjust = 0.9))
```

```{r plot-pgs-results, fig.asp=0.8}
pgs_test_res_df %>%
  filter(pgs_type == "gxe",
         set == "validation",
         thresh_name != "stacked") %>%
  left_join(optimal_pgs_df, by = c("e", "y", "pgs_type", "thresh_name")) %>%
  mutate(e = factor(e, levels = exposures, labels = exposures_clean),
         y = factor(y, levels = outcomes, labels = outcomes_clean),
         set = factor(set, levels = names(ukb_subsets), 
                      labels = paste0(str_to_title(names(ukb_subsets)), " set")),
         thresh_name = factor(thresh_name, 
                           levels = c(thresh_names, "stacked"),
                           labels = c(thresh_names_clean, "Stacked"))) %>%
  ggplot(aes(x = thresh_name, y = statistic)) +
  geom_bar(aes(color = optimal), stat = "identity") +
  scale_color_discrete(na.translate = FALSE) +
  facet_wrap(vars(e, y), ncol = 2, scale = "free_y", labeller = label_both) +
  labs(x = "", y = "Interaction test z-statistic",
       title = "iPGSxE optimization (validation set)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 0.9))

pgs_test_res_df %>%
  filter(set == "testing") %>%
  inner_join(optimal_pgs_df, by = c("e", "y", "pgs_type", "thresh_name")) %>%
  mutate(e = factor(e, levels = exposures, labels = exposures_clean),
         y = factor(y, levels = outcomes, labels = outcomes_clean),
         set = factor(set, levels = names(ukb_subsets), 
                      labels = paste0(str_to_title(names(ukb_subsets)), " set")),
         thresh_name = factor(thresh_name, 
                           levels = c(thresh_names, "stacked"),
                           labels = c(thresh_names_clean, "Stacked"))) %>%
  ggplot(aes(x = pgs_type, y = statistic, fill = pgs_type)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, color = "gray") +
  facet_wrap(vars(e, y), ncol = 2, scale = "free_y") +
  labs(x = "", y = "Interaction test z-statistic",
       title = "iPGSxE testing (testing set)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 0.9))
```

```{r example-viz}
e <- "whr"
y <- "hscrp_log"
tag <- "whr_hscrp_log"
thresh_name <- "thresh0.00001"
pgs_df <- read_csv(paste0("../data/processed/pgs/", tag, "_all_pgs.csv"), 
                   show_col_types = FALSE)
pgs_df <- select(pgs_df, id, pgs = {{thresh_name}})
regression_df <- inner_join(ukb_subsets$testing, pgs_df, by = "id") %>%
  mutate(pgs_bin = cut(pgs, quantile(pgs, seq(0, 1, length.out = 11)), 
                       include.lowest = TRUE, labels = paste0("Q", 1:10)))

regression_df %>%
  ggplot(aes(x = whr, y = hscrp_log, group = pgs_bin, color = pgs_bin)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(0.5, 1.5))

q1_corr <- cor.test(~ hscrp_log + whr, data = filter(regression_df, pgs_bin == "Q1"))$estimate
q10_corr <- cor.test(~ hscrp_log + whr, data = filter(regression_df, pgs_bin == "Q10"))$estimate

print(paste0("WHR-log(hsCRP) correlation in lowest vs. highest decile: ",
             round(q1_corr, 2), " versus ", round(q10_corr, 2)))
```

```{r exploration, eval = FALSE}
e <- "whr"
y <- "hscrp_log"
tag <- "whr_hscrp_log"
thresh_name <- "thresh0.01"
covars <- NULL
covars <- primary_covars
covars <- c(primary_covars, paste0("bmi", "By", gPCs))

pgs_df <- read_csv(paste0("../data/processed/pgs/", tag, "_all_pgs.csv"), 
                   show_col_types = FALSE)
pgs_df <- select(pgs_df, id, pgs = {{thresh_name}})

set <- "training"
regression_df <- inner_join(ukb_subsets[[set]], pgs_df, by = "id")
lm_form_str <- paste0(y, " ~ pgs * ", e)
if (!is.null(covars)) {
  lm_form_str <- paste0(lm_form_str, " + ", paste(covars, collapse = " + "))
}
lm_fit <- lm(as.formula(lm_form_str), data = regression_df %>% sample_n(70000))
lm_fit %>%
  broom::tidy() %>%
  filter(term == paste0("pgs:", e))
```

